---
title: "DS502- HW3"
author: "Mahdi Alouane and Rahul Pande"
output:
  pdf_document:
    # highlight: zenburn
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      tidy=TRUE,
                      fig.align='center',
                      tidy.opts=list(width.cutoff=60))
# https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html
```

### 1. (10 points) Section 6.8, page 259, question 2

(a) iii. holds true. Lasso puts a budget constraint on the parameters which decreases the model variance and it reduces overfitting. However, when we put a constraint, the model bias increases. From the bias-variance trade-off concept we can say that the lasso regression will give better prediction when its increase in bias is less that its decrease in variance.

(b) iii. holds true. Like Lasso above, Ridge also puts a budget constraint on the parameters which decreases the model variance and it reduces overfitting. However, when we put a constraint, the model bias increases. Similarly as above, from the bias-variance trade-off concept we can say that the Ridge regression will give better prediction when its increase in bias is less that its decrease in variance.

(c) ii. holds true. Since non-linear methods are more flexible, they have higher variance than least squares regression but lower bias. Again, from the bias variance trade-off, if the increase on model variance is less than the decrease in bias, then non-linear model will have better prediction accuracy.

```{r 6.8 2, echo=TRUE}
```

### 2. (20 points) Section 6.8, page264, question 11
### 3. (10 points) Section 7.9, Page 298, question 3
### 4. (10 points) Section 7.9, Page 298, question 4
### 5. (10 points) Section 7.9, Page 299, question 6
### 6. (20 points) Section 7.9, Page 299, question 7
### 7. (10 points) Section 8.4, Page 332, question 1
### 8. (10 points) Section 8.4, Page 333-334, question 8